Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-07-03 _tex/paper2.tex
Built site for gh-pages

However, Thurstone acknowledged that some assumptions of Case V are
problematic when researchers assess complex traits or heterogeneous
stimuli \citep{Thurstone_1927a}. Thus, given that contemporary CJ
applications often involve such traits and stimuli, two key assumptions
of Case V, and by extension, the BTL model, may not always hold in
theory or practice. These assumptions are the equal dispersion and zero
correlation between stimuli.

@@

Significant statistical and measurement issues can arise when the
assumption of equal dispersions between stimuli does not hold.
Specifically, the BTL model may overestimate the trait's reliability,
that is, the degree to which the outcome accurately reflects the
\emph{true} discriminal differences between stimuli. This
overestimation, in turn, results in spurious conclusions about these
differences \citep{McElreath_2020, Wu_et_al_2022} and, by extension,
about the underlying discriminal processes of stimuli.
Figure~\ref{fig-dispersion} also illustrates this scenario when the
model's discriminal difference distribution aligns with the thick
continuous line for \(\sigma_{B}-\sigma_{A}=0\), while the \emph{true}
discriminal difference follows any discontinuous line where
\(\sigma_{B}-\sigma_{A} \neq 0\). Furthermore, if researchers
acknowledge that \emph{misfit statistics} may represent texts with
different dispersions or outlying observations, the common CJ practice
of excluding stimuli based on these statistics may unintentionally
discard valuable information \citep{Miller_2023}, and introduce bias
into the trait estimates \citep{Zimmerman_1994, McElreath_2020}. The
direction and magnitude of these biases remain unpredictable, as they
depend on which stimuli researchers exclude from the analysis.

@@

Finally, similar to misfit stimuli, removing \emph{misfit judges} risks
discarding valuable information and introducing bias into trait
estimates \citep{Miller_2023}. The direction and magnitude of these
biases remain unpredictable, as they depend on which judges researchers
exclude from the analysis
\citep{Zimmerman_1994, OHagan_2018, McElreath_2020}. \emph{Misfit}
judges are those whose assessments deviate markedly from the shared
consensus
\citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018, Wu_et_al_2022},
often appearing as outliers under the BTL model \citep{Wu_et_al_2022}.

@@

Researchers in CJ studies typically use the BTL model to measure traits
and position the compared stimuli along a latent continuum
\citep{Thurstone_1927b}. The CJ literature shows that research
frequently relies on point estimates of these traits--typically the BTL
scores or its transformations--to conduct statistical inference or
hypothesis testing. For example, researchers have used these scores to
identify `misfit' judges and stimuli
\citep{Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018}, detect
biases in judges' ratings \citep{Pollitt_et_al_2003, Pollitt_2012b},
calculate correlations with other assessment methods
\citep{Goossens_et_al_2018, Bouwer_et_al_2023}, or test hypotheses
related to the underlying trait of interest
\citep{Casalicchio_et_al_2015, Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.

@@

The \emph{structural approach} to causal inference provides a formal
framework for identifying causes and estimating their effects using
data. The approach uses structural causal models (SCMs) and directed
acyclic graphs (DAGs)
\citep{Pearl_2009, Pearl_et_al_2016, Gross_et_al_2018, Neal_2020} to
formally and graphically represent the assumed causal structure of a
system, such as the one found in CJ assessments. Essentially, SCMs and
DAGs function as \emph{conceptual models} on which identification
analysis rests. \emph{Identification analysis} determines whether an
estimator can accurately compute an estimand based solely on its
(causal) assumptions, regardless of random variability
\citep{Schuessler_et_al_2023}. Here, \emph{estimands} represent the
specific quantities researchers aim to determine (i.e., a parameter)
\citep{Everitt_et_al_2010}. \emph{Estimators} denote the methods or
functions that transform data into an estimate (e.g., a statistical
model), while \emph{estimates} are the numerical values approximating
the estimand \citep{Neal_2020, Everitt_et_al_2010}.

A motivating example that will appear throughout this study clarifies
these concepts. In this example, researchers aim to answer the question:
``To what extent do different teaching methods influence students'
ability to produce high-quality written texts?'' To investigate this, a
researcher designs a CJ assessment by randomly assigning students
(individuals) to two groups, each receiving a different teaching method.
Judges then compare pairs of students' written texts (stimuli) to
produce a dichotomous outcome reflecting the relative quality of each
text (trait). Based on this setup, researchers can reformulate the
research question as the estimand: ``\emph{On average}, is there a
difference in the ability to produce high-quality written texts between
the two groups of students?''.

Following standard CJ practices, researchers would typically use
estimates from the BTL model--or its transformations--to approximate
this estimand. However, as discussed in Section~\ref{sec-theory-issues},
Thurstone's Case V and the BTL model exhibit several statistical and
measurement limitations. These limitations hinder the model's ability to
identify various estimands relevant to CJ inquiries, including the one
described in the example.

Fortunately, SCMs and DAGs support identification analysis through two
key advantages\footnote{In depth explanation of these topics is beyond
  the scope of this study, thus, readers seeking a more profound
  understanding can refer to introductory papers such as
  \citet{Pearl_2010}, \citet{Rohrer_2018}, \citet{Pearl_2019}, and
  \citet{Cinelli_et_al_2020}, and introductory books like
  \citet{Pearl_et_al_2018}, \citet{Neal_2020}, and
  \citet{McElreath_2020} are useful. For more advanced study, seminal
  papers such as \citet{Neyman_et_al_1923}, \citet{Rubin_1974},
  \citet{Spirtes_et_al_1991}, and \citet{Sekhon_2009}, along with books
  such as \citet{Pearl_2009}, \citet{Morgan_et_al_2014}, and
  \citet{Hernan_et_al_2025}, are recommended.}. First, regardless of
complexity, they can represent various causal structures using only five
fundamental building blocks \citep{Neal_2020, McElreath_2024}. This
feature allows researchers to decompose complex structures into
manageable components, facilitating their analysis. Second, they depict
causal relationships in a non-parametric way. This flexibility enables
feasible identification strategies without requiring specification of
the types of variables, the functional forms relating them, or the
parameters of those functional forms \citep{Pearl_et_al_2016}.

Thus, this section addresses the issues identified in
Section~\ref{sec-theory-issues} by extending Thurstone's general form
using the structural approach to causal inference. Specifically, it
combines Thurstone's core theoretical principles outlined in
Section~\ref{sec-thurstone_theory} with key CJ assessment design
features, such as the selection of judges, stimuli, and comparisons.
Section~\ref{sec-theory-theoretical_P} introduces the
\emph{conceptual-population model}, which incorporates these theoretical
principles and assumes an idealized setting where researchers observe a
\emph{conceptual population} of comparative judgment data--that is, data
representing all repeated judgments made by every available judge for
each pair of stimuli produced by each pair of individuals in the
population. Conversely, Section~\ref{sec-theory-theoretical_SC} presents
the \emph{sample-comparison model}, which integrates the assessment
design features and reflects a more realistic setting where researchers
access only a sample of judges, individuals, stimuli, and comparisons
from the conceptual population.

\subsection{The conceptual-population
model}\label{sec-theory-theoretical_P}

In the conceptual-population model, the idealized scenario of a
\emph{conceptual population} of comparative data enables the integration
of Thurstone's theoretical principles and provides a foundation for
proposing innovations aimed at addressing some of the issues discussed
in Section~\ref{sec-theory-issues}.

@@

SCM~\ref{fig-cj03_scm} presents the first theoretical principles
embedded in the conceptual-population model, which evaluates the impact
of different teaching methods on students' writing ability. This SCM
outlines the relationship between the conceptual-population outcome
\((O^{cp}_{iahbjk})\) and several related variables. The subscripts
\(i\) and \(h\) identify the students who authored the texts (i.e., the
individuals). The indices \(a\) and \(b\) represent the texts under
comparison (i.e., the stimuli). The index \(j\) indicates the judge
conducting the comparison, while the index \(k\) accounts for assessment
conditions where a judge compares the same pair of stimuli multiple
times, i.e., a \emph{repeated measures designs}
\citep[p.~366-376]{Lawson_2015}. Thus, the indexing system supports
comparisons between different texts written by the same student
\((i = h;\) \(a \neq b)\) and between texts written by distinct students
\((i \neq h;\) where \(a = b\) is permitted\()\), each compared once or
repeatedly by all judges \((j = 1,\dots,n_{J};\) \(k = 1,\dots,n_K;\)
where \(n_{J}>1\) and \(n_{K}\geq1)\). However, it excludes cases where
a judge compares a student's text to itself, whether once or multiple
times \((i = h;\) \(a = b;\) \(j = 1,\dots,n_{J};\)
\(k = 1,\dots,n_{K};\) where \(n_{J}>1\) and \(n_{K}\geq1)\), as such
comparison lacks practical relevance within the CJ framework. Here,
\(n_{J}\) indicates the total number of judges, and \(n_{K}\) denotes
the number of repeated judgments each judge performs.

@@

\caption{\label{fig-cj03}Conceptual-population model, scalar form.}

@@

\caption{\label{fig-cj04}Conceptual-population model, initial vectorized
form.}

@@

Building on the principles of Structural Equation Modeling (SEM)
\citep{Hoyle_et_al_2023} and Item Response Theory (IRT)
\citep{Fox_2010, vanderLinden_et_al_2017_I}, the conceptual-population
model integrates two \emph{hierarchical structural components} to
examine how different \emph{relevant} \footnote{\emph{Relevant
  variables} are those that satisfy the \emph{backdoor criterion}
  \citep[pp 37]{Neal_2020}, that is, they belong to a \emph{sufficient
  adjustment set}
  \citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}. A
  \emph{sufficient} set (potentially empty) blocks all non-causal paths
  between a predictor and an outcome without opening new ones
  \citep{Pearl_2009}. Refer also to footnote 1.} variables--whether
observed or latent--affect the primary latent variable of interest
\citep{Everitt_et_al_2010}. This hierarchical design enables researchers
to formulate and test hypotheses that account for both the nested
structure of stimuli and the uncertainties inherent in trait estimation
(see Section~\ref{sec-theory-issue1b} and
Section~\ref{sec-theory-issue2} for a discussion of these
considerations).

@@

Notably, all variables and functions shown in SCM~\ref{fig-cj09_scm} and
DAG~\ref{fig-cj09_dag} are part of the set of endogenous variables
\(V\), structural equations \(F\), and exogenous variables \(E\) for the
conceptual-population model. Additionally, the figures demonstrate that
all exogenous variables are independent of one another, as indicated by
the relationships \(e_{IA} \:\bot\:\{ e_{I}, e_{JK}, e_{J} \}\),
\(e_{I} \:\bot\:\{ e_{JK}, e_{J} \}\) and \(e_{JK} \:\bot\:e_{J}\) and
the absence of connecting arrows.

Overall, the conceptual-population model extends Thurstone's general
form by introducing key innovations to address the limitations discussed
in Section~\ref{sec-theory-issue1b} and Section~\ref{sec-theory-issue2}.
These enhancements include accounting for judges' biases and integrating
hierarchical structural components. Nevertheless, despite its promise of
enhancing measurement accuracy and precision, the model still depends on
the unrealistic assumption that researchers have access to data from the
\emph{conceptual population}. Since researchers rarely meet this
assumption in practice, they must consider a more realistic scenario.

@@

\caption{\label{fig-cj09}Conceptual-population model, final vectorized
form.}

@@

\subsection{The sample-comparison
model}\label{sec-theory-theoretical_SC}

The sample-comparison model presents a more realistic scenario than the
conceptual-population model. First, it explicitly assumes researchers
work with a data sample consisting of a limited number of repeated
judgments \((n^{s}_{K})\) from a sample of judges \((n^{s}_{J})\) and a
specific number of texts \((n^{s}_{A})\) from a sample of students
\((n^{s}_{I})\), all drawn from the conceptual population
(Section~\ref{sec-theory-theoretical_SC1}). Second, the model assumes
that judges do not perform \emph{all repeated judgments} within the data
sample (Section~\ref{sec-theory-theoretical_SC2}). Instead, they conduct
a sufficient number of stimuli comparisons, \(n_{C}\), to ensure an
accurate estimation of the proportion \(P(B>A)\), as proposed by
\citet{Thurstone_1927b}.

@@

Notably, Equation (\ref{eq-mat24}) shows that missing observations in
the vectors \(S_{I}\) and \(S_{J}\)--which represent unsampled students
and judges--directly determine which observations are missing in
\(S_{IA}\) and \(S_{JK}\). In other words, researchers can only observe
texts and judgments from students and judges initially included in the
sample. The equation also shows that the sum of observed elements in
\(S_{I}\) equals the number of sampled students \((n^{s}_{I})\) and that
a similar sum in vector \(S_{J}\) equals the sampled judges
\((n^{s}_{J})\). Conversely, the sum of observed elements in \(S_{IA}\)
represents the total sampled texts across all sampled students
\((n^{s}_{I} \cdot n^{s}_{A})\), while a similar sum in vector
\(S_{JK}\) represents the total sampled repeated judgments across all
sampled judges \((n^{s}_{J} \cdot n^{s}_{K})\). Notice that in this
example, because the design systematically excludes every third repeated
judgment, researchers can also express \(S_{JK}\) using
\(n_{K} = n^{s}_{K} = 2\).

@@

With the definition of \(S\), we incorporate the sample mechanism into
the conceptual-population model. Following the convention of
\citet{McElreath_2020} and \citet{Deffner_et_al_2022},
DAG~\ref{fig-cj14_dag} represents the conceptual-population outcome
\(O^{cp}_{R}\) as unobserved, emphasizing that researchers cannot
directly access this outcome due to the sampling mechanism. The DAG also
depicts the \emph{sample design} vector \(S\) as a causal factor
influencing the sample-comparison outcome \(O^{sc}_{R}\). A square
encloses \(S\), indicating that it is a conditioned variable. In this
context, \emph{conditioning} means that researchers restrict their focus
to the elements of \(O^{cp}_{R}\) that satisfy \(s_{(i,a,h,b,j,k)}=1\)
\citep{Neal_2020, McElreath_2020}. In essence, \(S\) is a vector that
selects \emph{all repeated judgments made by a subset of judges for a
subset of stimuli produced by the sampled individuals}.

@@

The DAG~\ref{fig-cj14_dag} also incorporates the \emph{comparison
mechanism} \(C\) into the conceptual-population model. It shows the
sample-comparison outcome \(O^{sc}_{R}\) as unobserved, emphasizing that
researchers cannot directly access this variable because of the
comparison mechanism. The DAG further shows \(C\) as a conditioned
variable (enclosed in a square) that causally influences the observed
outcome \(O_{R}\). This structure implies that \(C\) determines
\emph{which repeated judgments judges make for the stimuli produced by
the individuals}. In essence, \(C\) reflects the assumption that judges
\emph{do not} perform all possible repeated judgments but instead
complete a sufficient number, \(n_{C}\), to enable the accurate
estimation of the proportion \(P(B>A)\) for each stimulus pair
\citep[p.~267]{Thurstone_1927b}.

@@

In summary, the SCM~\ref{fig-cj15_scm} and DAG~\ref{fig-cj15_dag} extend
Thurstone's general form to address several limitations of the BTL
model. These extensions account for judge biases (see
Section~\ref{sec-theory-theoretical_P1}), reflect the hierarchical
structure of stimuli and incorporate measurement error in trait
estimation and hypothesis testing (see
Section~\ref{sec-theory-theoretical_P3}), and even clarify the role of
the sample and comparison mechanisms in CJ assessments (see
Section~\ref{sec-theory-theoretical_SC}). However, they do not resolve
concerns about the assumption of equal dispersions among stimuli
discussed in Section~\ref{sec-theory-issue1a}. Since this concern
relates to the statistical assumption underlying the distribution of the
discriminal process, we develop a formal statistical model to address it
in the next section.

@@

Using the structural causal model (SCM) \ref{fig-cj15_scm}, we can
derive a statistical model that addresses violations of the equal
dispersion assumption (see Section~\ref{sec-theory-issue1a}). This
derivation is possible because a fully specified SCM encodes functional
and probabilistic information, which we can replace with suitable
functions and probabilistic assumptions \citep{Pearl_et_al_2016}.
Specifically, SCM~\ref{fig-cj15_scm} allows us to express the joint
distribution of our complex CJ system as a product of simpler
conditional probability distributions (CPDs)\footnote{This re-expression
  is possible because the \emph{chain rule} of probability and the
  \emph{Bayesian Network Factorization (BNF)} property. For further
  details, see \citet{Pearl_et_al_2016} and \citet{Neal_2020}.}, as
shown in Equation (\ref{eq-mat51}). For clarity, we treat expressions
such as \(Y := f_{Y}(X)\), \(P(Y \mid X)\), and \(Y \sim f(Y \mid X)\)
as equivalent, where \(P(Y \mid X)\) and \(f(Y \mid X)\) represent the
CPD of \(Y\) given \(X\).
\begin{equation}\phantomsection\label{eq-mat51}{
\begin{aligned}
  P(O_{R}, & S, C, D_{R}, T_{IA}, X_{IA}, e_{IA}, T_{I}, X_{I}, e_{I}, B_{JK}, Z_{JK}, e_{JK}, B_{J}, Z_{J}, e_{J} ) & \\
  &= P(O_{R} \mid D_{R}, S, C) \cdot P(S) \cdot P(C) \cdot P(D_{R} \mid T_{IA}, B_{JK}) \\
  & \quad \cdot P(T_{IA} \mid T_{I}, X_{IA}, e_{IA}) \cdot P(T_{I} \mid X_{I}, e_{I}) \\
  & \quad \cdot P(B_{JK} \mid B_{J}, Z_{JK}, e_{JK}) \cdot P(B_{J} \mid Z_{J}, e_{J}) \\
  & \quad \cdot P(X_{IA}) \cdot P(X_{I}) \cdot P(Z_{JK}) \cdot P(Z_{J}) \\
  & \quad \cdot P(e_{IA}) \cdot P(e_{I}) \cdot P(e_{JK}) \cdot P(e_{J})  
\end{aligned}
}\end{equation}

>>>>

However, Thurstone acknowledged that some assumptions of Case V are
problematic when assessing complex traits or heterogeneous stimuli
\citep{Thurstone_1927a}. Thus, given that contemporary CJ applications
often involve such traits and stimuli, two key assumptions of Case V,
and by extension, the BTL model, may not always hold in theory or
practice. These assumptions are the equal dispersion and zero
correlation between stimuli.

@@

Significant statistical and measurement issues can arise when the
assumption of equal dispersions between stimuli does not hold.
Specifically, the BTL model may overestimate the trait's reliability,
that is, the degree to which the outcome accurately reflects the
\emph{true} discriminal differences between stimuli. This
overestimation, in turn, results in spurious conclusions about these
differences \citep{McElreath_2020, Wu_et_al_2022} and, by extension,
about the underlying discriminal processes of stimuli.
Figure~\ref{fig-dispersion} also illustrates this scenario when the
model's discriminal difference distribution aligns with the thick
continuous line for \(\sigma_{B}-\sigma_{A}=0\), while the \emph{true}
discriminal difference follows any discontinuous line where
\(\sigma_{B}-\sigma_{A} \neq 0\). Furthermore, if it is acknowledged
that \emph{misfit statistics} may represent texts with different
dispersions or outlying observations, the common CJ practice of
excluding stimuli based on these statistics may unintentionally discard
valuable information \citep{Miller_2023}, and introduce bias into the
trait estimates \citep{Zimmerman_1994, McElreath_2020}. The direction
and magnitude of these biases remain unpredictable, as they depend on
which stimuli are excluded from the analysis.

@@

Finally, similar to misfit stimuli, removing \emph{misfit judges} risks
discarding valuable information and introducing bias into trait
estimates \citep{Miller_2023}. The direction and magnitude of these
biases remain unpredictable, as they depend on which judges are excluded
from the analysis \citep{Zimmerman_1994, OHagan_2018, McElreath_2020}.
\emph{Misfit} judges are those whose assessments deviate markedly from
the shared consensus
\citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018, Wu_et_al_2022},
often appearing as outliers under the BTL model \citep{Wu_et_al_2022}.

@@

In CJ studies, the BTL model is typically used to measure traits and
position the compared stimuli along a latent continuum
\citep{Thurstone_1927b}. The CJ literature shows that studies frequently
relies on point estimates of these traits--typically the BTL scores or
its transformations--to conduct statistical inference or hypothesis
testing. For example, researchers have used these scores to identify
`misfit' judges and stimuli
\citep{Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018}, detect
biases in judges' ratings \citep{Pollitt_et_al_2003, Pollitt_2012b},
calculate correlations with other assessment methods
\citep{Goossens_et_al_2018, Bouwer_et_al_2023}, or test hypotheses
related to the underlying trait of interest
\citep{Casalicchio_et_al_2015, Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.

@@

The \emph{structural approach} to causal inference
\citep{Pearl_2009, Pearl_et_al_2016} offers a formal framework for
identifying causes and estimating their effects using data. The approach
relies on structural causal models (SCMs) and directed acyclic graphs
(DAGs) to formally and graphically represent the assumed causal
structure of a system
\citep{Morgan_et_al_2014, Gross_et_al_2018, Neal_2020}, such as the one
found in CJ assessments. In essence, SCMs and DAGs function as
\emph{conceptual models} on which identification analysis rests
\citep{Schuessler_et_al_2023}. \emph{Identification analysis} determines
whether an estimator can accurately compute an estimand based solely on
its (causal) assumptions, regardless of random variability. Here,
\emph{estimands} represent the specific quantities researchers aim to
determine (i.e., a parameter) \citep{Everitt_et_al_2010}.
\emph{Estimators} denote the methods or functions that transform data
into an estimate (e.g., a statistical model), while \emph{estimates} are
the numerical values approximating the estimand
\citep{Neal_2020, Everitt_et_al_2010}.

As an illustration, consider a researcher seeking to answer the
question: ``To what extent do different teaching methods influence
students' ability to produce high-quality written texts?'' To
investigate this, the researcher designs a CJ assessment by randomly
assigning students (individuals) to two groups, each exposed to a
different teaching method. Judges then compare pairs of students'
written texts (stimuli) to produce a dichotomous outcome reflecting the
relative quality of each text (trait). Based on this setup, researchers
can reformulate the research question as the estimand: \emph{On average,
is there a difference in the ability to produce high-quality written
texts between the two groups of students?} Following standard CJ
practices, the researcher would then use point estimates from the BTL
model, or its transformations, to approximate this estimand. However, as
discussed in Section~\ref{sec-theory-issue1}, Thurstone's Case V and the
BTL model exhibit several statistical and measurement limitations. These
limitations hinder the model's ability to identify and accurately
estimate a range of estimands relevant to CJ inquiries, including the
one described in this illustration.

Fortunately, SCMs and DAGs support identification analysis through two
key advantages\footnote{In depth explanation of these topics is beyond
  the scope of this study, thus, readers seeking a more profound
  understanding can refer to introductory papers such as
  \citet{Pearl_2010}, \citet{Rohrer_2018}, \citet{Pearl_2019}, and
  \citet{Cinelli_et_al_2020}, and introductory books like
  \citet{Pearl_et_al_2018}, \citet{Neal_2020}, and
  \citet{McElreath_2020} are useful. For more advanced study, seminal
  papers such as \citet{Neyman_et_al_1923}, \citet{Rubin_1974},
  \citet{Spirtes_et_al_1991}, and \citet{Sekhon_2009}, along with books
  such as \citet{Pearl_2009}, \citet{Morgan_et_al_2014}, and
  \citet{Hernan_et_al_2025}, are recommended.}. First, regardless of
complexity, they can represent various causal structures using only five
fundamental building blocks \citep{Neal_2020, McElreath_2024}. This
feature allows the decomposition of complex structures into manageable
components, facilitating their analysis. Second, they depict causal
relationships in a non-parametric way. This flexibility enables feasible
identification strategies without requiring specification of the types
of variables, the functional forms relating them, or the parameters of
those functional forms \citep{Pearl_et_al_2016}.

Thus, using SCMs and DAGs, this study extends Thurstone's general form
to address the issues identified in Section~\ref{sec-theory-issues}.
This extension combines Thurstone's core theoretical principles (see
Section~\ref{sec-thurstone_theory}) with key CJ assessment design
features, such as the selection of judges, stimuli, and comparisons.
Specifically, Section~\ref{sec-theory-theoretical_P} introduces the
\emph{conceptual-population model} (henceforth CPM), which incorporates
these theoretical principles and assumes an idealized setting where
researchers observe a \emph{conceptual population} of comparative
judgment data--that is, data representing all repeated judgments made by
every available judge for each pair of stimuli produced by each pair of
individuals in the population. Conversely,
Section~\ref{sec-theory-theoretical_SC} presents the
\emph{sample-comparison model} (hereafter SCM), which integrates the
assessment design features and reflects a more realistic setting where
researchers access only a sample of judges, individuals, stimuli, and
comparisons from the conceptual population.

\subsection{The conceptual-population model
(CPM)}\label{sec-theory-theoretical_P}

In the CPM, the idealized scenario of a \emph{conceptual population} of
comparative data enables the integration of Thurstone's theoretical
principles and provides a foundation for proposing innovations aimed at
addressing some of the issues discussed in
Section~\ref{sec-theory-issues}.

@@

SCM~\ref{fig-cj03_scm} presents the first theoretical principles
embedded in the CPM evaluating the impact of different teaching methods
on students' writing ability. This SCM outlines the relationship between
the conceptual-population outcome \((O^{cp}_{iahbjk})\) and several
related variables. The subscripts \(i\) and \(h\) identify the students
who authored the texts (i.e., the individuals). The indices \(a\) and
\(b\) represent the texts under comparison (i.e., the stimuli). The
index \(j\) indicates the judge conducting the comparison, while the
index \(k\) accounts for assessment conditions where a judge compares
the same pair of stimuli multiple times, i.e., a \emph{repeated measures
designs} \citep{Lawson_2015}. Thus, the indexing system supports
comparisons between different texts written by the same student
\((i = h;\) \(a \neq b)\) and between texts written by distinct students
\((i \neq h;\) where \(a = b\) is permitted\()\), each compared once or
repeatedly by all judges \((j = 1,\dots,n_{J};\) \(k = 1,\dots,n_K;\)
where \(n_{J}>1\) and \(n_{K}\geq1)\). However, it excludes cases where
a judge compares a student's text to itself, whether once or multiple
times \((i = h;\) \(a = b;\) \(j = 1,\dots,n_{J};\)
\(k = 1,\dots,n_{K};\) where \(n_{J}>1\) and \(n_{K}\geq1)\), as such
comparison lacks practical relevance within the CJ framework. Here,
\(n_{J}\) indicates the total number of judges, and \(n_{K}\) denotes
the number of repeated judgments each judge performs.

@@

\caption{\label{fig-cj03}Conceptual-population model (CPM), scalar
form.}

@@

\caption{\label{fig-cj04}Conceptual-population model (CPM), initial
vectorized form.}

@@

Building on the principles of Structural Equation Modeling (SEM)
\citep{Hoyle_et_al_2023} and Item Response Theory (IRT)
\citep{Fox_2010, vanderLinden_et_al_2017_I}, the CPM integrates two
\emph{hierarchical structural components} to examine how different
\emph{relevant} \footnote{\emph{Relevant variables} are those that
  satisfy the \emph{backdoor criterion} \citep[pp 37]{Neal_2020}, that
  is, they belong to a \emph{sufficient adjustment set}
  \citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}. A
  \emph{sufficient} set (potentially empty) blocks all non-causal paths
  between a predictor and an outcome without opening new ones
  \citep{Pearl_2009}. Refer also to footnote 1.} variables--whether
observed or latent--affect the primary latent variable of interest
\citep{Everitt_et_al_2010}. This hierarchical design supports the
formulation and testing of hypotheses that account for both the nested
structure of stimuli and the uncertainties inherent in trait estimation
(see Section~\ref{sec-theory-issue1b} and
Section~\ref{sec-theory-issue2} for a discussion of these
considerations).

@@

Notably, all variables and functions shown in SCM~\ref{fig-cj09_scm} and
DAG~\ref{fig-cj09_dag} are part of the set of endogenous variables
\(V\), structural equations \(F\), and exogenous variables \(E\) for the
CPM. Additionally, the figures demonstrate that all exogenous variables
are independent of one another, as indicated by the relationships
\(e_{IA} \:\bot\:\{ e_{I}, e_{JK}, e_{J} \}\),
\(e_{I} \:\bot\:\{ e_{JK}, e_{J} \}\) and \(e_{JK} \:\bot\:e_{J}\) and
the absence of connecting arrows.

Overall, the CPM extends Thurstone's general form by introducing key
innovations to address the limitations discussed in
Section~\ref{sec-theory-issue1b} and Section~\ref{sec-theory-issue2}.
These enhancements include accounting for judges' biases and integrating
hierarchical structural components. Nevertheless, despite its promise of
enhancing measurement accuracy and precision, the model still depends on
the unrealistic assumption that researchers have access to data from the
\emph{conceptual population}. Since this assumption is rarely met in
practice, the study must consider a more realistic scenario.

@@

\caption{\label{fig-cj09}Conceptual-population model (CPM), final
vectorized form.}

@@

\subsection{The sample-comparison model
(SCM)}\label{sec-theory-theoretical_SC}

The SCM presents a more realistic scenario than the CPM. First, it
explicitly assumes a data sample consisting of a limited number of
repeated judgments \((n^{s}_{K})\) from a sample of judges
\((n^{s}_{J})\) and a specific number of texts \((n^{s}_{A})\) from a
sample of students \((n^{s}_{I})\), all drawn from the conceptual
population (Section~\ref{sec-theory-theoretical_SC1}). Second, the model
assumes that judges do not perform \emph{all repeated judgments} within
the data sample (Section~\ref{sec-theory-theoretical_SC2}). Instead,
they conduct a sufficient number of stimuli comparisons, \(n_{C}\), to
ensure an accurate estimation of the proportion \(P(B>A)\), as proposed
by \citet{Thurstone_1927b}.

@@

Notably, Equation (\ref{eq-mat24}) shows that missing observations in
the vectors \(S_{I}\) and \(S_{J}\)--which represent unsampled students
and judges--directly determine which observations are missing in
\(S_{IA}\) and \(S_{JK}\). In other words, researchers can only observe
texts and judgments from students and judges initially included in the
sample. The equation also shows that the sum of observed elements in
\(S_{I}\) equals the number of sampled students \((n^{s}_{I})\) and that
a similar sum in vector \(S_{J}\) equals the sampled judges
\((n^{s}_{J})\). Conversely, the sum of observed elements in \(S_{IA}\)
represents the total sampled texts across all sampled students
\((n^{s}_{I} \cdot n^{s}_{A})\), while a similar sum in vector
\(S_{JK}\) represents the total sampled repeated judgments across all
sampled judges \((n^{s}_{J} \cdot n^{s}_{K})\). Notice that in this
example, because the design systematically excludes every third repeated
judgment, \(S_{JK}\) can also be expressed using
\(n_{K} = n^{s}_{K} = 2\).

@@

With the definition of \(S\), we incorporate the sample mechanism into
the CPM. Following the convention of \citet{McElreath_2020} and
\citet{Deffner_et_al_2022}, DAG~\ref{fig-cj14_dag} represents the
conceptual-population outcome \(O^{cp}_{R}\) as unobserved, emphasizing
that this outcome cannot be directly accessed due to the sampling
mechanism. The DAG also depicts the \emph{sample design} vector \(S\) as
a causal factor influencing the sample-comparison outcome
\(O^{sc}_{R}\). A square encloses \(S\), indicating that it is a
conditioned variable. In this context, \emph{conditioning} means that
the analysis is restricted to the elements of \(O^{cp}_{R}\) that
satisfy \(s_{(i,a,h,b,j,k)}=1\) \citep{Neal_2020, McElreath_2020}. In
essence, \(S\) is a vector that selects \emph{all repeated judgments
made by a subset of judges for a subset of stimuli produced by the
sampled individuals}.

@@

The DAG~\ref{fig-cj14_dag} also incorporates the \emph{comparison
mechanism} \(C\) into the CPM. It shows the sample-comparison outcome
\(O^{sc}_{R}\) as unobserved, emphasizing that this variable cannot be
directly accessed because of the comparison mechanism. The DAG further
shows \(C\) as a conditioned variable (enclosed in a square) that
causally influences the observed outcome \(O_{R}\). This structure
implies that \(C\) determines \emph{which repeated judgments judges make
for the stimuli produced by the individuals}. In essence, \(C\) reflects
the assumption that judges \emph{do not} perform all possible repeated
judgments but instead complete a sufficient number, \(n_{C}\), to enable
the accurate estimation of the proportion \(P(B>A)\) for each stimulus
pair \citep[p.~267]{Thurstone_1927b}.

@@

In summary, the SCM~\ref{fig-cj15_scm} and DAG~\ref{fig-cj15_dag} extend
Thurstone's general form to address several limitations of the BTL
model. These extensions account for judges' biases (see
Section~\ref{sec-theory-theoretical_P1}), reflect the hierarchical
structure of stimuli and incorporate measurement error in trait
estimation and hypothesis testing (see
Section~\ref{sec-theory-theoretical_P3}), and even clarify the role of
the sample and comparison mechanisms in CJ assessments (see
Section~\ref{sec-theory-theoretical_SC}). However, they do not resolve
concerns about the assumption of equal dispersions among stimuli
discussed in Section~\ref{sec-theory-issue1a}. Since this concern
relates to the statistical assumption underlying the distribution of the
discriminal process, we develop a formal statistical model to address it
in the next section.

@@

Using the SCM~\ref{fig-cj15_scm}, we can derive a statistical model that
addresses violations of the equal dispersion assumption (see
Section~\ref{sec-theory-issue1a}). This derivation is possible because a
fully specified SCM encodes functional and probabilistic information,
which we can replace with suitable functions and probabilistic
assumptions \citep{Pearl_et_al_2016}. Specifically,
SCM~\ref{fig-cj15_scm} allows us to express the joint distribution of
our complex CJ system as a product of simpler conditional probability
distributions (CPDs)\footnote{This re-expression is possible because the
  \emph{chain rule} of probability and the \emph{Bayesian Network
  Factorization (BNF)} property. For further details, see
  \citet{Pearl_et_al_2016} and \citet{Neal_2020}.}, as shown in Equation
(\ref{eq-mat51}). For clarity, we treat expressions such as
\(Y := f_{Y}(X)\), \(P(Y \mid X)\), and \(Y \sim f(Y \mid X)\) as
equivalent, where \(P(Y \mid X)\) and \(f(Y \mid X)\) represent the CPD
of \(Y\) given \(X\). \begin{equation}\phantomsection\label{eq-mat51}{
\begin{aligned}
  P(O_{R}, & S, C, D_{R}, T_{IA}, X_{IA}, e_{IA}, T_{I}, X_{I}, e_{I}, B_{JK}, Z_{JK}, e_{JK}, B_{J}, Z_{J}, e_{J} ) & \\
  &= P(O_{R} \mid D_{R}, S, C) \cdot P(S) \cdot P(C) \cdot P(D_{R} \mid T_{IA}, B_{JK}) \\
  & \quad \cdot P(T_{IA} \mid T_{I}, X_{IA}, e_{IA}) \cdot P(T_{I} \mid X_{I}, e_{I}) \\
  & \quad \cdot P(B_{JK} \mid B_{J}, Z_{JK}, e_{JK}) \cdot P(B_{J} \mid Z_{J}, e_{J}) \\
  & \quad \cdot P(X_{IA}) \cdot P(X_{I}) \cdot P(Z_{JK}) \cdot P(Z_{J}) \\
  & \quad \cdot P(e_{IA}) \cdot P(e_{I}) \cdot P(e_{JK}) \cdot P(e_{J})  
\end{aligned}
}\end{equation}
